Comparing variability across fMRI analysis software packages and numerical errors

### 2021-12-13 paper/paper.tex

\newcommand{\fslspm}{FSL-SPM\xspace}
\newcommand{\fslafni}{FSL-AFNI\xspace}
\newcommand{\afnispm}{AFNI-SPM\xspace}
\newcommand{\tristan}[1]{\color{orange}\textbf{From Tristan:} #1\color{black}\xspace}
\newcommand{\ali}[2]{\color{green}\textbf{Ali:} #1\color{black}\xspace}
\newcommand{\discuss}[1]{\uwave{#1}}
\newcommand{\gk}[1]{\color{purple}#1 \textbf{-GK}\color{black}\xspace}

@@

\maketitle
\begin{abstract}

Variability has been broadly observed in functional MRI analyses across both
differences in analytic tools and numerical instabilities. However,
the relationship between numerical (within-tool) and between-tool variabilities for
given analysis is unclear. We extended a previous comparison
of fMRI analysis software libraries (namely FSL, AFNI, and SPM) and relate the observed differences to the numerical
variability observed using Monte-Carlo arithmetic. We found that
between-tool variability was consistently larger than numerical variability in group-level analyses,
whereas numerical variability approached between-tool variability in some subject-level analyses.
Interestingly, between-tool variability and
machine error appeared moderately correlated ($r > 0.5$) in all tested conditions.
Finally, we found that numerical variability had effects of similar magnitude to between-tool
variability when errors were introduced at the precision of 17 bits --- a
precision level found in commonly-used libraries \gk{What does this mean? They only have precision of 17, or they use data that is 17 bits?}. Our findings motivate the
continued investigation of numerical instability in neuroimaging, and position it 
as a possible proxy for studying arbitrary between-tool variations.
\end{abstract}

@@

Software variability has been highlighted as an important source of error
in fMRI analysis, originating from mainly two factors: between-tool
variability denotes differences in the outcomes of different
software toolboxes --- such as FSL, AFNI and SPM --- when analyzing the same dataset
using similar methods \gk{this sentence says "two factors" but then only mentioned between tool. Also, it is too long}. Numerical variability, in contrast, denotes
differences in outcomes of the same analytical toolbox
executed in different environments – such as different operating systems or
different hardware platforms \gk{you mention these two examples of numerical variability, but you use neither in this paper. Why not mention the one you use?}. This manuscript investigates the extent to
which both types of variability relate to each other.

Numerical variability stems from errors introduced in floating point
computations by numerical rounding, cancellation, and absorption~\cite{muller2018handbook} \gk{only? What about poor conditioning + random effects + noise in data?}. In
practice, the occurrence of such errors is affected by variations in
hardware, software dependencies, compilation, parallelization, and most
likely other factors \gk{what does "most likely other factors" mean?}. In the best case, the impact of such errors and the induced
variations is in the range of 1 unit in the last place (ulp) --- also known
as machine error, the difference between two successive floating-point
numbers. However, when computations depend on imprecise libraries,
numerical error may vary by several ulps. For instance, the numerical error
of the complementary error function (erfc) is larger than 3 ulps in several
mathematical libraries~\cite{zimmermann:hal-03141101}, implying that an update of this function could introduce
numerical perturbations larger than 3 ulps. Due to numerical instability,
machine error may amplify during its propagation through analyses,
resulting in considerable loss of precision in final results. In fMRI analysis,
numerical variability produced by Linux updates has been shown to result in Dice
coefficients ranging from 0.0 to 1.0 for ICA components produced by FSL
MELODIC~\cite{Glatard2015} \gk{there isn't really context for what this result means, yet. Is this a lot? Dice hasn't been explained. Also, how are ICA components (weighted vectors) compared via Dice (a measure for binary image/vector comparison)? }.

Between-tool variability is not a surprising concept: different programs
executed on the same data may obviously produce different results \gk{even when doing conceptually the same thing? Why? Why is this obvious?}. However,
in fMRI analysis, programs all aim to estimate brain activity from measured
variations in the BOLD signal, which is not expected to be measurably
impacted by computational artifacts \gk{Why wouldn't it be? Especially if you state that it isn't surprising for tools to lead to different results, how come you expect fMRI to not suffer from this?}. Therefore, the observation reported
in~\cite{bowring2019exploring} that the three main fMRI analysis toolboxes
may result in marked analytical differences (Dice coefficients ranging from
0.000 to 0.684 for thresholded activation maps), has raised substantial
concern. In a follow-up study~\cite{bowring2021isolating}, the same authors
isolated the analytical components where the pipelines diverge,
highlighting a major impact of the signal model and suggesting that the
root-cause for between-tool variability depends on the analysis design and
task paradigm \gk{If you state this finding, you need to unpack what it means a little bit. What part of analysis design? By task paradigm do you essentially mean the expected effect-size for the task that individuals are performing within the scanner?}. Further, the work in~\cite{Li2021.12.01.470790} found
moderate agreement between results produced by five independent fMRI
analysis pipelines and analyzed in details the components that led to these
variations \gk{This sentence feels inconsistent and counter to the previous, and also doesn't tell us their findings; what components led to variations?}. Between-tool variability is supposedly an important determinant
of the substantial analytical variability recently observed in
fMRI~\cite{botvinik2020variability} \gk{arguably, one of the larger issues that the NARPS result showed us was that people construct and test ``equivalent'' hypotheses differently, so despite highly similar data they will arrive at different conclusions}.

@@

%%%%%%%%%% Summary of statstics %%%%%%%%
\setlength{\tabcolsep}{4pt}
\begin{table}[h]
    \centering
    \begin{tabular}{|c|l|c|c|c|}
        \hline
%        \multirow{2}{*}{} & \multicolumn{1}{c}{Thresholded}& & \multicolumn{1}{c}{Unthresholded}& \\
        \multicolumn{2}{|c|}{} & FSL & AFNI & SPM \\
        \hline
        {Preprocessing} & {Motion Correction}                          & \checkmark    & \checkmark     & \checkmark  \\
        {} & {Segmentation}                               &    &      & \checkmark  \\
        {} & {Brain Extraction (Anatomical)}              & \checkmark     & \checkmark    & \checkmark  \\
        {} & {Brain Extraction (Functional)}              &   & \checkmark     &  \\
        {} & {Intra-subject Coregistration}               & \checkmark    & \checkmark     & \checkmark \\
        {} & {Inter-subject Registration}                 & \checkmark    & \checkmark     & \checkmark \\
        {} & {Analysis Voxel Size}                        & \checkmark    & \checkmark     & \checkmark \\
        {} & {Smoothing}                                  & \checkmark    & \checkmark     & \checkmark  \\
        \hline
        {First-level} & {Model Specification}                          & \checkmark    & \checkmark     & \checkmark  \\
        {} & {Inclusion of 6 Motion Parameters}                               & \checkmark   &  \checkmark    & \checkmark  \\
        {} & {Model Estimation}                           & &     & \checkmark  \\
        {} & {Contrasts}                                   &  \checkmark & \checkmark     & \checkmark \\
        \hline
        {Second-level} & {Model Specification}                          & \checkmark    & \checkmark     & \checkmark  \\
        {} & {Model Estimation}                           &      &    & \checkmark  \\
        {} & {Contrasts}                                   &   & \checkmark     & \checkmark  \\
        {} & {Second-level Inference}                               &  \checkmark  &    \checkmark  & \checkmark  \\
        \hline

      \end{tabular}
    \caption{Software processing steps (adapted from~\cite{bowring2019exploring}).}
    \label{table:pipeline-steps}
\end{table}

@@

To introduce numerical noise in the analyses, we used
Fuzzy Libmath~\cite{salari2021accurate}, a version of the GNU
mathematical library (libmath) instrumented with Monte-Carlo arithmetic.
Monte-Carlo arithmetic simulates numerical errors
errors by introducing a controlled amount of noise in floating-point
operations through the following perturbation~\cite{Parker1997-qq}:

@@

We measured numerical (within-tool; WT) variability by running the same analyses three
times using Fuzzy Libmath with a virtual precision of $t=53$~bits for
double-precision values and $t=24$~bits for single-precision values. These
values were chosen such that the numerical perturbation simulates machine
error. The resulting samples were equally plausible estimates of
the true numerical result at the precision used by the pipelines. Moreover, to evaluate numerical perturbations at different magnitudes, 
we repeated the FSL analyses for virtual
precisions ranging from $t=1$~bit to $t=24$~bits for single-precision
and double-precision values. We also repeated the FSL analyses for virtual 
precisions ranging from $t=24$~bits to $t=53$~bits for double-precision values, having 
set the virtual precision to $t=24$~bits for single-precision values.

@@

Further, from the thresholded maps, we determined regional instability
between activation clusters in the 360 regions in the Human Connectome
Project Multi-Modal Parcellation atlas version 1.0
(HCP-MMP1.0)~\cite{glasser2016multi}. For BT, we considered a region
unstable for a pair of tools if it contained activated voxels for a tool
but not for the other one. For WT, we considered a region unstable for a
pair of tools (A, B) if for tool A or tool B it contained activated voxels only for some Fuzzy Libmath
samples.

@@

\setlength{\tabcolsep}{5pt}
\begin{table}[h]
    \centering
    \begin{tabular}{cccccc|cc}
        \toprule
        && \multicolumn{4}{c|}{Group map} & \multicolumn{2}{c}{Subject maps}\\
        \multirow{2}{*}{}& {} & \multicolumn{2}{c}{Thresholded} & \multicolumn{2}{c|}{Unthresholded} & \multicolumn{2}{c}{Unthresholded} \\
        % \cmidrule{3-8}
        {} & {} & $\mu$ & $\sigma$ & $\mu$ & $\sigma$ & $\mu$ & $\sigma$ \\
        \midrule
        \rowcolor{lightgray!50}
        {Between Tools} & FSL vs. SPM        &  1.282       & 0.525      & 0.443     & 0.344  & 0.366       & 0.293     \\
        \rowcolor{lightgray!50}
        {(BT)} & FSL vs. AFNI                &  1.548       & 0.616      & 0.547     & 0.441  & 0.439       & 0.352     \\
        \rowcolor{lightgray!50} 
        {} & AFNI vs. SPM                    &  1.475       & 0.672      & 0.608     & 0.477  & 0.491       & 0.381     \\
        {Within Tool} & FSL                  &  0.354       & 0.491      & 0.082     & 0.065  & 0.077       & 0.054     \\
        {(WT)}   & SPM                       &  0.252       & 0.448      & 0.054     & 0.045  & 0.048       & 0.037     \\
        {}   & AFNI                          &  0.434       & 0.524      & 0.128     & 0.135  & 0.108       & 0.131     \\
        \bottomrule
    \end{tabular}
    \caption{Voxel-wise mean and standard deviation of BT and WT variability
    in t-statistics maps.}
    \label{table:pipeline-stats}
\end{table}

@@

The finding that numerical variability approached tool variability at the
virtual precision of t=17~bits is interesting too. Indeed, while machine
error generally introduces differences in the order of 1 ulp --- or
t=53~bits for double-precision values and t=24~bits for single-precision
ones --- common scientific software dependencies introduce larger errors.
For instance, SciPy's 2D spline interpolation was recently found to be
precise up to t=10~bits~\cite{pytracer} \tristan{Fix ref} and might therefore introduce
numerical perturbations leading to errors in the range of between-tool
variability \tristan{Yohan, please check that sentence}. Such high errors
might be triggered by updates in operating systems, Python, MATLAB, and
other software dependencies.

>>>>

\newcommand{\fslspm}{FSL-SPM\xspace}
\newcommand{\fslafni}{FSL-AFNI\xspace}
\newcommand{\afnispm}{AFNI-SPM\xspace}
\newcommand{\tristan}[1]{\color{orange}\textbf{From Tristan:} #1\color{black}\xspace}
\newcommand{\ali}[2]{\color{green}\textbf{Ali:} #1\color{black}\xspace}
\newcommand{\discuss}[1]{\uwave{#1}}
\newcommand{\gk}[1]{\color{purple}#1 \textbf{-GK}\color{black}\xspace}
\newcommand{\yohan}[1]{\color{cyan!75!black} \textbf{Yohan:} #1\color{black}\xspace}
\newcommand{\yohanmod}[1]{\color{cyan!75!black} \sout{#1}\color{black}\xspace}

@@

\maketitle
\begin{abstract}
  Variability has been broadly observed in functional MRI analyses across both
  differences in analytic tools and numerical instabilities. However,
  the relationship between numerical (within-tool) and between-tool variabilities for
  given analysis is unclear. We extended a previous comparison
  of fMRI analysis software libraries (namely FSL, AFNI, and SPM) and relate the observed differences to the numerical
  variability observed using Monte-Carlo arithmetic. We found that
  between-tool variability was consistently larger than numerical variability in group-level analyses,
  whereas numerical variability approached between-tool variability in some subject-level analyses.
  Interestingly, between-tool variability and
  machine error appeared moderately correlated ($r > 0.5$) in all tested conditions.
  Finally, we found that numerical variability had effects of similar magnitude to between-tool
  variability when errors were introduced at the precision of 17 bits --- a
  precision level found in commonly-used libraries \gk{What does this mean? They only have precision of 17, or they use data that is 17 bits?}. \yohan{agree with Greg, it is unclear and a strong assertion!} Our findings motivate the
  continued investigation of numerical instability in neuroimaging, and position it
  as a possible proxy for studying arbitrary between-tool variations.
\end{abstract}

@@

Software variability has been highlighted as an important source of error
in fMRI analysis, originating from mainly two factors: between-tool
variability denotes differences in the outcomes of different
software toolboxes --- such as FSL, AFNI and SPM --- when analyzing the same dataset
using similar methods \gk{this sentence says "two factors" but then only mentioned between tool. Also, it is too long}. \yohan{What do you mean by methods? Numerical schemes? If so, are we sure that theses tools are using the same numerical schemes?} Numerical variability, in contrast, denotes
differences in outcomes of the same analytical toolbox
executed in different environments – such as different operating systems or
different hardware platforms \gk{you mention these two examples of numerical variability, but you use neither in this paper. Why not mention the one you use?}. This manuscript investigates the extent to
which both types of variability relate to each other.

Numerical variability stems from errors introduced in floating point
computations by \yohanmod{numerical} rounding, cancellation, and absorption~\cite{muller2018handbook} \gk{only? What about poor conditioning + random effects + noise in data?}. In
practice, the occurrence of such errors is affected by variations in
hardware, software dependencies, compilation, parallelization, and most
likely other factors \gk{what does "most likely other factors" mean?}. In the best case, the impact of such errors and the induced
variations is in the range of \yohanmod{1} \yohan{1/2 ulp for round-to-nearest} unit in the last place (ulp) --- also known
as machine error, the difference between two successive floating-point
numbers. However, when computations depend on imprecise libraries,
numerical error may vary by several ulps. For instance, the \yohan{maximal} numerical error
of the complementary error function (erfc) is larger than 3 ulps in several
mathematical libraries~\cite{zimmermann:hal-03141101}, implying that an update of this function could introduce
numerical \yohanmod{perturbations} \yohan{differences} larger than 3 ulps. Due to numerical instability,
machine error may amplify during its propagation through analyses \yohan{do you mean calculations?},
resulting in considerable loss of precision in final results. In fMRI analysis,
numerical variability produced by Linux updates has been shown to result in Dice
coefficients ranging from 0.0 to 1.0 for ICA components produced by FSL
MELODIC~\cite{Glatard2015} \gk{there isn't really context for what this result means, yet. Is this a lot? Dice hasn't been explained. Also, how are ICA components (weighted vectors) compared via Dice (a measure for binary image/vector comparison)? }.

Between-tool variability is not a surprising concept: different programs
executed on the same data may obviously produce different results \gk{even when doing conceptually the same thing? Why? Why is this obvious?}. However,
in fMRI analysis, programs all aim to estimate brain activity from measured
variations in the BOLD signal, which is not expected to be measurably
impacted by computational artifacts \gk{Why wouldn't it be? Especially if you state that it isn't surprising for tools to lead to different results, how come you expect fMRI to not suffer from this?}. \yohan{+1} Therefore, the observation reported
in~\cite{bowring2019exploring} that the three main fMRI analysis toolboxes
may result in marked analytical differences (Dice coefficients ranging from
0.000 to 0.684 for thresholded activation maps), has raised substantial
concern. In a follow-up study~\cite{bowring2021isolating}, the same authors
isolated the analytical components where the pipelines diverge,
highlighting a major impact of the signal model and suggesting that the
root-cause for between-tool variability depends on the analysis design and
task paradigm \gk{If you state this finding, you need to unpack what it means a little bit. What part of analysis design? By task paradigm do you essentially mean the expected effect-size for the task that individuals are performing within the scanner?}. Further, the work in~\cite{Li2021.12.01.470790} found
moderate agreement between results produced by five independent fMRI
analysis pipelines and analyzed in details the components that led to these
variations \gk{This sentence feels inconsistent and counter to the previous, and also doesn't tell us their findings; what components led to variations?}. Between-tool variability is supposedly an important determinant
of the substantial analytical variability recently observed in
fMRI~\cite{botvinik2020variability} \gk{arguably, one of the larger issues that the NARPS result showed us was that people construct and test ``equivalent'' hypotheses differently, so despite highly similar data they will arrive at different conclusions}.

@@

%%%%%%%%%% Summary of statstics %%%%%%%%
\setlength{\tabcolsep}{4pt}
\begin{table}[h]
  \centering
  \begin{tabular}{|c|l|c|c|c|}
    \hline
    %        \multirow{2}{*}{} & \multicolumn{1}{c}{Thresholded}& & \multicolumn{1}{c}{Unthresholded}& \\
    \multicolumn{2}{|c|}{} & FSL                                & AFNI       & SPM                     \\
    \hline
    {Preprocessing}        & {Motion Correction}                & \checkmark & \checkmark & \checkmark \\
    {}                     & {Segmentation}                     &            &            & \checkmark \\
    {}                     & {Brain Extraction (Anatomical)}    & \checkmark & \checkmark & \checkmark \\
    {}                     & {Brain Extraction (Functional)}    &            & \checkmark &            \\
    {}                     & {Intra-subject Coregistration}     & \checkmark & \checkmark & \checkmark \\
    {}                     & {Inter-subject Registration}       & \checkmark & \checkmark & \checkmark \\
    {}                     & {Analysis Voxel Size}              & \checkmark & \checkmark & \checkmark \\
    {}                     & {Smoothing}                        & \checkmark & \checkmark & \checkmark \\
    \hline
    {First-level}          & {Model Specification}              & \checkmark & \checkmark & \checkmark \\
    {}                     & {Inclusion of 6 Motion Parameters} & \checkmark & \checkmark & \checkmark \\
    {}                     & {Model Estimation}                 &            &            & \checkmark \\
    {}                     & {Contrasts}                        & \checkmark & \checkmark & \checkmark \\
    \hline
    {Second-level}         & {Model Specification}              & \checkmark & \checkmark & \checkmark \\
    {}                     & {Model Estimation}                 &            &            & \checkmark \\
    {}                     & {Contrasts}                        &            & \checkmark & \checkmark \\
    {}                     & {Second-level Inference}           & \checkmark & \checkmark & \checkmark \\
    \hline
  \end{tabular}
  \caption{Software processing steps (adapted from~\cite{bowring2019exploring}).}
  \label{table:pipeline-steps}
\end{table}

@@

To introduce numerical noise in the analyses, we used
Fuzzy Libmath~\cite{salari2021accurate}, a version of the GNU
mathematical library (libmath) instrumented with Monte-Carlo arithmetic.
Monte-Carlo arithmetic simulates numerical errors
by introducing a controlled amount of noise in floating-point
operations through the following perturbation~\cite{Parker1997-qq}:

@@

We measured numerical (within-tool; WT) variability by running the same analyses three
times using Fuzzy Libmath with a virtual precision of $t=53$~bits for
double-precision values and $t=24$~bits for single-precision values. These
values were chosen such that the numerical perturbation simulates machine
error. The resulting samples were equally plausible estimates of
the true \yohan{How can you assert that when you don't have ground truth? Would it be more safe to say "plausible estimates numerical solutions"?} numerical result at the precision used by the pipelines. Moreover, to evaluate numerical perturbations at different magnitudes,
we repeated the FSL analyses for virtual
precisions ranging from $t=1$~bit to $t=24$~bits for single-precision
and double-precision values. We also repeated the FSL analyses for virtual
precisions ranging from $t=24$~bits to $t=53$~bits for double-precision values, having
set the virtual precision to $t=24$~bits for single-precision values.

@@

Further, from the thresholded maps, we determined regional instability
between activation clusters in the 360 regions in the Human Connectome
Project Multi-Modal Parcellation atlas version 1.0
(HCP-MMP1.0)~\cite{glasser2016multi}. For BT, we considered a region
unstable for a pair of tools if it contained activated voxels for a tool
but not for the other one. For WT, we considered a region unstable for a
pair of tools (A, B) if for tool A or tool B it contained activated voxels only for some Fuzzy Libmath
samples. \yohan{A voxel is activated if its intensity is $>$ 0? If so, the given definition focuses on regions with low intensities while regions with high intensities may be more unstable}

@@

\setlength{\tabcolsep}{5pt}
\begin{table}[h]
  \centering
  \begin{tabular}{cccccc|cc}
    \toprule
                      &              & \multicolumn{4}{c|}{Group map}  & \multicolumn{2}{c}{Subject maps}                                                                     \\
    \multirow{2}{*}{} & {}           & \multicolumn{2}{c}{Thresholded} & \multicolumn{2}{c|}{Unthresholded} & \multicolumn{2}{c}{Unthresholded}                               \\
    % \cmidrule{3-8}
    {}                & {}           & $\mu$                           & $\sigma$                           & $\mu$                             & $\sigma$ & $\mu$ & $\sigma$ \\
    \midrule
    \rowcolor{lightgray!50}
    {Between Tools}   & FSL vs. SPM  & 1.282                           & 0.525                              & 0.443                             & 0.344    & 0.366 & 0.293    \\
    \rowcolor{lightgray!50}
    {(BT)}            & FSL vs. AFNI & 1.548                           & 0.616                              & 0.547                             & 0.441    & 0.439 & 0.352    \\
    \rowcolor{lightgray!50}
    {}                & AFNI vs. SPM & 1.475                           & 0.672                              & 0.608                             & 0.477    & 0.491 & 0.381    \\
    {Within Tool}     & FSL          & 0.354                           & 0.491                              & 0.082                             & 0.065    & 0.077 & 0.054    \\
    {(WT)}            & SPM          & 0.252                           & 0.448                              & 0.054                             & 0.045    & 0.048 & 0.037    \\
    {}                & AFNI         & 0.434                           & 0.524                              & 0.128                             & 0.135    & 0.108 & 0.131    \\
    \bottomrule
  \end{tabular}
  \caption{Voxel-wise mean and standard deviation of BT and WT variability
    in t-statistics maps.}
  \label{table:pipeline-stats}
\end{table}

@@

The finding that numerical variability approached tool variability at the
virtual precision of t=17~bits is interesting too. Indeed, while machine
error generally introduces differences in the order of \yohanmod{1} \yohan{1/2} ulp
\yohan{Are we talking about one arithmetic operation? Or an entire execution?}
--- or
t=53~bits for double-precision values and t=24~bits for single-precision
ones \yohan{I would not mix virtual precision and error measurement}--- common scientific software dependencies introduce larger errors
.
For instance, SciPy's 2D spline interpolation was recently found to be
precise up to t=10~bits~\cite{pytracer} \tristan{Fix ref}
\yohan{Same as previous, I would talk about significance or relative error, or remove the t= and not talk about virtual precision} and might therefore introduce
numerical perturbations leading to errors in the range of between-tool
variability \tristan{Yohan, please check that sentence}. Such high errors
might be triggered by updates in operating systems, Python, MATLAB, and
other software dependencies.
\yohan{The sentence is ok, just a picky point: the SciPy result you cited describes the precision of the spline implementation while at the end you're talking differences due to updates.
  The nuance is that two versions may have a similar precision and so when you udpate your
  system, the difference between both versions is low but each version can be ill-conditionned.}
